---
title: "Basic Visualization"
author: "Sumad Singh"
date: "September 7, 2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### BASIC GRAPHIC SYSTEM IN R
### PARAMETERS TO ANNOTATE GRAPHS
### COMBINE GRAPHS

```{r}
## USING BASIC GRAPHIC SYSTEM IN R
#1.LAYERING 
#We want to build layers of plots on a single chart, R provides a system to do that
#help("dev.cur")
# IDEA IS TO OPEN A GRAPHICS DEVICE, R ALLOWS TO OPEN MULTIPLE DEVICES,
# DIRECT OUTPUT TO A CHOICE OF DEVICE
# FUNCTIONS LIKE PDF(), PNG(), JPEG() OPEN A DEVICE BY THEMSELVES
# HIGH LEVEL FUNCTIONS LIKE PLOT(), HIST() ETC ALSO OPEN A DEVICE

# USEFUL FUNCTIONS FOR OPENING AND MANAGING DEVICES
# a Null device is always open in a session, once a new device is opened Null does not show up
#dev.new() # Open a new device, 
# dev.cur() # lists the name and number of current active device
# dev.list() # lists the names and mumbers of all open devices
# dev.off() # shuts off a device
# dev.next() # shuffle to next device w.r.t the current
# dev.previous() # same, goes to previous
# graphics.off()  # closes all open devices

# EXAMPLE OF PLOTTING A SCATTER PLOT, AND A REGRESSION LINE
attach(mtcars)
dev.new()
plot(x = hp ,y = mpg,type = "p") # PLOT IS A HIGH LEVEL FUNCTION IN R, WHICH CAN MAKE DIFFERENT TYPE OF PLOTS
# BASED ON THE 'TYPE' ARGUMENT
abline(reg = lm(mpg~hp))
detach(mtcars)
#help(abline)

```

```{r}
#2. KEY PLOTTING FUNCTIONS
# A. PLOT
#PLOT FUNCTION WORKS DIFFERENTY ON OBJECTS LIKE DATA FRAMES, DENSITY, PLOT.DEFAULT WORKS 
#FOR SCATTERPLOT B/W X AND Y
attach(mtcars)
plot(x= hp, y= mpg,type = "p",xlab = "log (horsepower)",ylab = "log(Miles per gallon)",log = "xy" ,main = " Log Linear relationship?"  ,
     sub =  "log mpg vs log hp" ,frame.plot = TRUE,asp = 1)

detach(mtcars)
# xlim, ylim are limits on the axes
# main title
# sub - subtitle of the plot
# xlab/ylab -  x , y axis labels
# axes - used "xaxt" or "yaxt to suppresss one of thre axes"
# frame.plot - logical to choose if plot to be enclosed in a box
# asp - aspect ration i.e y/x

```
```{r}
# Plot function operates on a data frame with all numeric columns to give
# bivariate scatter plots
plot(mtcars)
```
```{r}
#3. GRAPHICAL PARAMETERS - COLOR, LINE/POINT/BAR TYPE, LENGTH,WIDTH ,TEXT ETC
#A. GENERAL FUNCTION TO SPECIFY GRAPHICAL PARAMETERS
#par() # return a list of current graphical paramters
par.default <- par(no.readonly = TRUE) # return the current setting of parameters that can be modified, use of no.readonly=T
par(lty = 2, lwd = 2, pch= 2 ,cex= 1.2) # new settings
attach(mtcars)
plot(x= hp, y= mpg,type = "p",xlab = "horsepower",ylab = "Miles per gallon" ,main = " Linear relationship?"  ,
     sub =  "mpg vs hp" ,frame.plot = TRUE)
abline(reg = lm(mpg ~ hp))
detach(mtcars)
par(par.default) # return to original settings


#B. GRAPHICAL PARAMETERS WITHIN HIGH LEVEL FUNCTIONS (NOT ALL ARE AVAILABLE THOUGH!, 
#SO USE A OPTION AS NEEDED )
# plot() and abline() for example have ... argument to specify graphical parameters,
# these settings are used only for the current plot
```

### GRAPHICAL PARAMETERS SUMMARY

|GEOM|PARAMETER/AESTHETIC|KEYWORD USED  FOR SETTING AESTHETIC FOR GEOM|REALIZATIONS -<AES.GEOM> |
|---|---|---|
|TEXT - TITLE, SUBTITLE,AXIS LABELS,AXIS VALUES|FONT FAMILY,FONT,SIZE,COLOUR|family,font,cex,col|family, 
font.main, font.sub, font.lab, font.axis AND col.main, col.sub etc AND cex.main, cex.sub etc
|
|POINT|SIZE, TYPE|
|LINE| TYPE, WIDTH|
```{r}
#C.  PARAMETERS FOR POINT AND LINES
# pch -  point type, goes from 0 to 24
# cex - size zooming relative to default of 1, expressed as 1.<x>
# lty - line type goes from 1 to 6
# lwd - line width ,elative to default of 1, expressed as 1.<x>

# D. PARAMETERS FOR COLORS
# col - color of lines or points in the plot, can be given as a vector, the values of which are recycled
# col.axis - color of axes
# col.lab - color of label of axes
# col.main , col.sub 
#fg, bg - foreground and background color

# FUNCTIONS TO GET A VECTOR OF COLORS, COLORS CAN BE SPECIFIED AS NAMES, OR DIGITS
#OR HEXADECIMALS
x <- colors(5)
y <- rainbow(5)
z <- heat.colors(5)
cvec <- c(rgb(0.2,1,0.5), rgb(0.5,1,0),rgb(0.5,0.5,0.2),rgb(0.1,0.3,0.4),rgb(0.4,.2,0.1)) 
# rgb lets you create  vector of colors using red green blue
hvec <- c(hsv(0.2,1,0.5), hsv(0.5,1,0),hsv(0.5,0.5,0.2),hsv(0.1,0.3,0.4),hsv(0.4,.2,0.1))
# hsv is another function
pie(x = c(100,20,30,40,10),radius = 1 ,clockwise = TRUE ,col = y ,labels = y)
# function to create a pie chart
```
```{r}
#E. PARAMETERS FOR TEXT -
#SIZE : params for font size
# cex - specify how much text enlargement is needed for plotted text
# same paramter govern point size
# cex. axis - for text on the axis
# cex.lab - axes labels
# cex.main - main title
# cex.sub - subtitle


#FONT FAMILY, STYLE
# family - for drawing text,
  #by default there are three families - serif, sans, mono
# font - integer for specifying bold  -2, italic-3 ,plain - 1, bold italic - 4  
 # for plotted text
# font. axis - for text on the axis, i.e values marked on axes
# font.lab - axes labels
# font.main - main title
# font.sub - subtitle

# ps - font point size (1/72 inch). text size =  cex * ps


#F. GRAPH DIMESNIONS AND MARGINS
#pin(width, height) to fix plot dimensios in inches
#mar() - vector of 4 to fix margins in lines in order (bottom, left, top, right) 
#mai() - same but this is in inches, default  is c(5,4,4,2) + 0.1 

# Example , draw a simple scatter and line chart 
# fix dim of plot, fix margins, note the margins are taken from the plot are and out
attach(mtcars)
par.default <- par(no.readonly = TRUE)
par(pin=c(5,3), mai = c(1,1,0.5,0.5)) # 33% of y dim as margin in bottom, 20% on left

# fix point type/size,
# Font family, font bolded for title, italics for axis, bold italics for labels
plot(x = wt, y= mpg, type = "p" ,main = "Scatter Plot with regression line" ,
     frame.plot = TRUE, pch = 2,col= "blue",xlab = "Weight", ylab = "Mpg",
     font.axis = 3, font.main = 2,font.lab = 4, 
     cex = 1.2, cex.main = 1.2, cex.lab= 0.8, cex.axis = 0.9)

# fix color/type/width of line
abline(reg=lm(mpg~wt), lty = 2, lwd = 1.2, col ="red")
detach(mtcars)
par(par.default)
```

```{r}
# G. ADDING TEXT ON CHART, LEGENDS, CUSTOMIZING AXES SEPARATELY
# Only some of the high level functionslike plot() let you specify
# the graphical parameters, in other cases separate function can be used

#G.1 TITLE LETS TEXT SIZE, FONT, COLOR
#title(main = ,sub = ,xlab = ,ylab = , # Adding titles - main, sub, labels
#      col.main =,col.sub, col.lab =,  # Adjust color
#      cex.main =, cex.sub, cex.lab)  # Adjust size

  
```

